---
title: "Maize-Phyllachora maydis RNAseq"
author: "Emily Roggenkamp"
date: "2024-05-01"
output: html_document
---


```{r}
rm(list=ls()) 
setwd("C:/Users/emrog/OneDrive - Michigan State University/Chilvers/Tar spot/Corn resistance/Maize_RNAseq") 
```

Input Files from HTseq count output

```{r}
myFiles <- list.files(pattern=glob2rx("*htseq.tsv"), # Create a vector containing the names of your files
                      all.files=T, 
                      full.names=F)

myFiles 

myData <- lapply(myFiles, # Use list apply to apply the native R function read.table() to myFiles.
                 read.table, colClasses=c("character","numeric"))

names(myData) <- c("early_control_rep1", # Name the elements of myData.
                   "early_control_rep2",
                   "early_control_rep3",
                   "early_infected_rep1",
                   "early_infected_rep2",
                   "early_infected_rep3",
                   "late_control_rep1", 
                   "late_control_rep2",
                   "late_control_rep3",
                   "late_infected_rep1", 
                   "late_infected_rep2",
                   "late_infected_rep3")

```

Remove bottom summary lines. This is a really quick and dirty solution. This is a function to select the first 44,303 lines.

```{r}
rmLines <- function(x){ 
  x <- x[1:44303,]
}

myData <- lapply(myData, rmLines) # Apply the rmLines() function to each object in the myData list.

nrow(myData$early_control_rep2) # Check to see how many lines there are in each element of myData now. There should be only 42,189.
```

Generate combined data frame for counts data

```{r}
counts <- data.frame(row.names=myData$early_control_rep1[,1],
                      early_control_rep1=myData$early_control_rep1[,2],
                      early_control_rep2=myData$early_control_rep2[,2],
                      early_control_rep3=myData$early_control_rep3[,2],
                      early_infected_rep1=myData$early_infected_rep1[,2],
                      early_infected_rep2=myData$early_infected_rep2[,2],
                      early_infected_rep3=myData$early_infected_rep3[,2],
                      late_control_rep1=myData$late_control_rep1[,2],
                      late_control_rep2=myData$late_control_rep2[,2],
                      late_control_rep3=myData$late_control_rep3[,2],
                      late_infected_rep1=myData$late_infected_rep1[,2],
                      late_infected_rep2=myData$late_infected_rep2[,2],
                      late_infected_rep3=myData$late_infected_rep3[,2]
)

head(counts)
#write.table(counts, file="raw_counts.txt", quote=F, sep="\t")
```

```{r}
geneTotals <- rowSums(counts) # Evaluate the sum of each row and save to a new vector called geneTotals.

countsNonZero <- counts[geneTotals>0,] # Subset the rows where the geneTotal is greater than 0.

nrow(countsNonZero) # See how many genes are left in the analysis.
treatments <- data.frame(c(rep("10day_unexposed", 3), rep("10day_exposed", 3), 
                           rep("24day_unexposed", 3), rep("24day_exposed", 3)))
treatments$treatment <- as.factor(c("unexposed","unexposed","unexposed","exposed","exposed","exposed",
                          "unexposed","unexposed","unexposed","exposed","exposed","exposed"))
treatments$time <- as.factor(c("10day","10day","10day","10day","10day","10day",
                               "24day","24day","24day","24day","24day","24day"))
names(treatments) <- c("sample", "treatment", "time")
rownames(treatments) <- c("early_control_rep1",
                   "early_control_rep2",
                   "early_control_rep3",
                   "early_infected_rep1",
                   "early_infected_rep2",
                   "early_infected_rep3",
                   "late_control_rep1", 
                   "late_control_rep2",
                   "late_control_rep3",
                   "late_infected_rep1", 
                   "late_infected_rep2",
                   "late_infected_rep3")

# We will define the 6 columns of our data set according to their treatment type, the factor in this experiment.
```

Install and Load packages

```{r, echo = FALSE, results = "hide"}
#install.packages("BiocManager")
#BiocManager::install(c("limma","edgeR","DESeq2","EnhancedVolcano","DEGreport"))
library(DEGreport)
library(EnhancedVolcano)
library(apeglm)
library(limma) # load the packages we need next
library(edgeR)
library(DESeq2)
library(tidyverse)
library(dplyr)
library(tidyr)
library(ggplot2)
```


```{r}
colData <- DataFrame(treatments) # treatments was our column labels organized as factors
```

Create the DESeq object from the counts matrix we made that has no non-expressed genes, uses the new colData object we made to label the tissues, and the same experimental design using treatments as the nominal categories.

This Model allows me to make the contrasts I want from the same object. Still trying to figure out if there's a way to do this in a formula with both "conditions" like below. Also using these thresholds, I get a lot less DEGs.

Found making complicated model was difficult and didn't provide contrasts based on treatment at each time separately. Combining the treatment_time conditions into the sample factor allowed me to make those contrasts.

```{r}
dds <- DESeqDataSetFromMatrix(countsNonZero, 
                              colData=colData, 
                              design=formula(~sample))
```

Pull out normalized counts for later
```{r, Normalization}
dds_norm <- estimateSizeFactors(dds)
sizeFactors(dds_norm)
normalized_counts <- counts(dds_norm, normalized=TRUE)
```

```{r}
dds <- DESeq(dds)
res_all <- results(dds)
res_all <- res_all[!(is.na(res_all$padj)),]
sigGenesadj <- as.data.frame(res_all[res_all$padj < 0.05, ])
```

```{r}
res10 <- results(dds, contrast = c("sample", "10day_exposed", "10day_unexposed"), 
                 alpha = 0.05, lfcThreshold = 1)
res24 <- results(dds, contrast = c("sample", "24day_exposed", "24day_unexposed"), 
                 alpha = 0.05, lfcThreshold = 1)
res10 <- res10[!(is.na(res10$padj)),]
res24 <- res24[!(is.na(res24$padj)),]
sigGenes10 <- as.data.frame(res10[res10$padj < 0.05, ])
sigGenes24 <- as.data.frame(res24[res24$padj < 0.05, ])
sigGenes10$ensembl_gene_id <- rownames(sigGenes10) 
sigGenes24$ensembl_gene_id <- rownames(sigGenes24)
sigGenes10 <- subset(sigGenes10, select = c(7,1,2,3,4,5,6))
sigGenes24 <- subset(sigGenes24, select = c(7,1,2,3,4,5,6))
#write.table(sigGenes10, file="DEG_10dpe_0501.txt", quote=F, sep="\t", col.names=NA)
#write.table(sigGenes24, file="DEG_24dpe_0501.txt", quote=F, sep="\t", col.names=NA)
```

# Figure 2A

Look at numbers of DEGs up and down regulated at each timepoint and put into a bar graph

```{r}
summary(res10)
summary(res24)
```

```{r}
groupA <- tibble(up = 999, down = 2161)
groupB <- tibble(up = 1543, down = 2410)

all <- bind_rows(
  mutate(groupA, group = "10dpe"),
  mutate(groupB, group = "24dpe")
) %>% 
  pivot_longer(up:down)
```

```{r}
deg <- 
ggplot(all, aes(x = value, y = group, fill = name)) +
  geom_col(position = position_dodge()) +
  geom_text(aes(label=value), vjust=1.6, position = position_dodge(0.9), color="white", size=3.5) +
  theme_classic() +
  xlab("Number of DEGs") +
  ylab(NULL) +
  guides(fill = guide_legend(reverse=TRUE)) +
  scale_fill_manual(labels = c("Down-Regulated", "Up-Regulated"), values = c("#39558CFF","#74D055FF")) +
  theme(axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10),
        axis.text = element_text(size = 10),
        legend.title = element_blank(),
        legend.position = "bottom") +
  coord_flip()

deg

#ggsave(plot = deg, filename = "deg2.png", units = "mm", width = 78, height = 78, dpi = 600)
```

# Figure 2B

Ouput DEGs at each timepoint and input into online portal to find overlap.

```{r}
#deg_overlap <- merge(sigGenes10, sigGenes24, by.x = "gene", by.y = "gene")
#deg_overlap <- deg_overlap[,1:23]
#write.csv(deg_overlap, file="DEG_both.csv")
```

# PCA Plot - Supplemental Figure 1A

```{r}
tdds <- DESeqTransform(dds)
PCA <- plotPCA(tdds, intgroup = c("sample")) +
  geom_point(aes(size=4)) +
    scale_colour_manual(name = "Sample Type", values = c("#6a51a3",
                                "#4292c6",
                                "#238b45",
                                "#ff7f00")) +
  #scale_color_viridis(discrete=TRUE) if wanting to use viridis palettes
    theme_classic() +
    theme(plot.title = element_text(size = 20, face = "bold", hjust = 0.5)) +
    theme(axis.title = element_text(angle = 0, size = 12, face = "bold")) +
    theme(axis.text.x = element_text(angle = 0, size = 12,hjust = 0.5, vjust = 0.5)) +
    theme(axis.text.y = element_text(angle = 0, size = 12, hjust = 0.5, vjust = 0.5)) +
    theme(legend.title = element_text(size = 12, face = "bold"), 
          legend.text = element_text(size = 12)) +
  guides(fill = "none", color = guide_legend(nrow = 4, override.aes = list(size=5)), size = "none")
PCA
```

Variance Stabilization

```{r}
p_mean_sd_scaled <- 
  countsNonZero %>% 
  as.data.frame() %>% 
  rownames_to_column("gene") %>% 
  pivot_longer(cols = - gene, names_to = "sample", values_to = "counts") %>% 
  group_by(gene) %>% 
  summarise(gene_average = mean(counts), gene_stdev = sd(counts)) %>% 
  ungroup() %>% 
  ggplot(., aes(x = log10(gene_average), y = log10(gene_stdev))) +
  geom_point(alpha = 0.5, fill = "grey", colour = "black") +
  labs(x = "Gene count average (log10 scale)",
       y = "Gene count standard deviation (log10 scale)") +
  ggtitle("Mean - Standard deviation relationship\n(no variance stabilisation ")
p_mean_sd_scaled
```

```{r}
vsd <- varianceStabilizingTransformation(object = dds, 
                                               blind = TRUE,           # do not take the design formula into account
                                               fitType = "parametric")
variance_stabilised_counts <- assay(vsd)
p_mean_sd_vst <- 
  variance_stabilised_counts %>% 
  as.data.frame() %>% 
  rownames_to_column("gene") %>% 
  pivot_longer(cols = - gene, names_to = "sample", values_to = "counts") %>% 
  group_by(gene) %>% 
  summarise(gene_average = mean(counts), gene_stdev = sd(counts)) %>% 
  ungroup() %>% 
  ggplot(., aes(x = gene_average, y = gene_stdev)) +
  geom_point(alpha = 0.5, fill = "grey", colour = "black") +
  labs(x = "Gene count average (variance stabilised)", 
       y = "Gene count standard deviation (variance stabilised)") +
  ggtitle("Mean - Standard deviation relationship\n(after variance stabilisation ")
p_mean_sd_vst

```

Dispersion

```{r}
geneDispersion <- dds@rowRanges@elementMetadata@listData$dispGeneEst
treatmentsx <- dds@rowRanges@elementMetadata@listData$sample_10dpe_infected_vs_10dpe_control

plot(treatmentsx, # See how fold change corresponds to dispersion.
     geneDispersion, 
     xlab="Fold Change Expression", ylab="Gene Dispersion Estimate", pch=19, cex=0.4)

# Want low dispersion and low fold change

plotDispEsts(dds) # Check the trend of normalized counts to see how it corresponds to the dispersion estimate.
```
# Volcano Plots

10 dpe

```{r}
resLFC <- lfcShrink(dds = dds,
                  res = res10,
                  type = "apeglm",
                  coef = "sample_10day_exposed_vs_10day_unexposed")

EnhancedVolcano(resLFC,
                x = "log2FoldChange",
                y = "padj",
                lab = rownames(resLFC),
                xlim = c(-15, +15),
                ylim = c(0,150),
                pCutoff = 1e-08,
                FCcutoff = 4, 
                title = "Phyllachora maydis 10dpe"
                )

```

24 dpe

```{r}
resLFC2 <- lfcShrink(dds = dds,
                  res = res24,
                  type = "apeglm",
                  coef = "sample_24day_exposed_vs_24day_unexposed")

EnhancedVolcano(resLFC2,
                x = "log2FoldChange",
                y = "padj",
                lab = rownames(resLFC),
                xlim = c(-15, +15),
                ylim = c(0,150),
                pCutoff = 1e-08,
                FCcutoff = 4, 
                title = "Phyllachora maydis 24dpe"
                )

```

# Supplemental Figure 1B

Pearson's correlation coefficients of log fold change.

```{r}
corResult <- cor(t(as.matrix(dds@assays@data$counts[res_all$padj < 0.01,]))) 
corr_coeff <- cor(dds@assays@data$counts, method = "pearson")

Pearsons1 <- as.matrix(corr_coeff) %>%
pheatmap::pheatmap(., 
                   color = viridis(50, direction = -1),
                   main = "Pearson correlation",
                   cluster_rows = T, cluster_cols = T,
                   labels_row = c("10day unexposed-1","10day unexposed-2",
                                  "10day unexposed-3",
                                  "10dpe-1","10dpe-2","10dpe-3",
                                  "24day unexposed-1","24day unexposed-2",
                                  "24day unexposed-3",
                                  "24dpe-1","24dpe-2","24dpe-3"),
                   show_colnames = FALSE,
                   display_numbers = round(corr_coeff, 2), number_color = "white"
                   )
```


